### 2018-03-15 paper/4-build.tex
Lots more from Simon

The focus of this paper is on a variety of implementations of \hs{Build}, given
a client-supplied implementation of \hs{Task}. That is, we are going to take
\hs{Task} as given from now on, and explore \hs{Build} variants: first
abstractly (in this section) and then concretely in~\S\ref{sec-implementations}.

Let \hs{result}~\hs{=}~\hs{build}~\hs{task}~\hs{key}~\hs{store}. The build
system \hs{build} is responsible for ensuring that \hs{key} and its dependencies
are up-to-date in \hs{result} starting from \hs{store} (see the formal definition
of correctness in~\S\ref{sec-build-correctness}). As part of that work,
\hs{build} \emph{may} store auxiliary information \hs{info} of type \hs{i} in
the \hs{store}, which will be provided to future calls of \hs{build}.

All the build systems we define are polymorphic in both \hs{k} and \hs{v}
(perhaps adding constraints to them), and have specific instantiations of \hs{c}
and \hs{i}. As an example, recall that the \hs{busy} build system defined
in~\S\ref{sec-general-build}, has \hs{c}~\hs{=}~\hs{Monad}, i.e. it supports
monadic tasks, and \hs{i = ()}, i.e. it does not store any build information.

@@

The topological approach computes a linear order, which when followed, ensures
the \hs{build} is correct regardless of the initial \hs{store}. Given a function
from a key to its dependencies, and the output \hs{key}, you can compute the
linear order by first taking the transitive closure of \hs{key}'s dependencies,
and then computing its topological sort. However, as we have seen
in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task,
which requires the build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling
out dynamic dependencies.

@@

The topological approach has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time -- while the
actions themselves may be incremental the pre-processing is not. An alternative
approach is to store the topological order between build runs and assume it to
be correct, but if the build determines it is wrong, reorder.

This requires a way to abort tasks that have failed due to out-of-date
dependencies. It is also not minimal in the sense that tasks start, may do some
meaningful work, then abort. However, in the case of an \hs{Applicative} system,
that work is zero.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}) is to simply build dependencies when they are
requested. By combining that with a transient store of which keys have already
been built, you can obtain a minimal build system.

@@

\subsection{Determining out-of-date keys}

@@

The idea of a dirty bit is to have one piece of information, whether a key is
\emph{dirty} or \emph{clean}. After a build, all bits are set to clean. When the
next build starts anything that changed between the two states is marked
dirty~--~by marking additional things dirty/clean the build system can track
what needs to rebuild. When reaching a key, if it and all its dependencies are
clean, the key does not need recomputing.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell. Since \Excel is the only manner by which a cell may
be changed by the user, it can mark cells dirty as necessary. When rebuilding,
if a cell only depends on clean cells it is skipped, otherwise it is rebuilt and
marked dirty itself.

@@

\Make uses file modification times, and compares files to their dependencies,
which can be thought of as a dirty bit which is set when a file is newer than
its dependencies. The interesting property of this dirty bit is that it is not
under control of \Make, but an existing bit that is repurposed. In particular,
rebuilding a file automatically clears its dirty bit, and automatically sets
the dirty bit of the nodes depending on it. One thing \Make does require is that
file timestamps only go forward in time -- something that can be violated by
backup software.

@@

\subsection{Summary}

>>>>

The focus of this paper is on a variety of implementations of \hs{Build}, given
a \emph{client-supplied} implementation of \hs{Task}. That is, we are going to take
\hs{Task} as given from now on, and explore variants of \hs{Build}: first
abstractly (in this section) and then concretely in~\S\ref{sec-implementations}.

\todo{SLPJ}{What does this para add?}
Let \hs{result}~\hs{=}~\hs{build}~\hs{task}~\hs{key}~\hs{store}. The build
system \hs{build} is responsible for ensuring that \hs{key} and its dependencies
are up-to-date in \hs{result} starting from \hs{store} (\S\ref{sec-build-correctness}). As part of that work,
\hs{build} \emph{may} store auxiliary information \hs{info} of type \hs{i} in
the \hs{store}, which will be provided to future calls of \hs{build}.

\todo{SLPJ}{What does this para add?}
All the build systems we define are polymorphic in both \hs{k} and \hs{v}
(perhaps adding constraints to them), and have specific instantiations of \hs{c}
and \hs{i}. As an example, recall that the \hs{busy} build system defined
in~\S\ref{sec-general-build}, has \hs{c}~\hs{=}~\hs{Monad}, i.e. it supports
monadic tasks, and \hs{i = ()}, i.e. it does not store any build information.

@@

The topological approach pre-computes a linear order, which when followed, ensures
the \hs{build} is correct regardless of the initial \hs{store}. Given a function
from a key to its dependencies, and the output \hs{key}, you can compute the
linear order by first taking the transitive closure of \hs{key}'s dependencies,
and then computing its topological sort. However, as we have seen
in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task,
which requires the build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling
out dynamic dependencies.

@@

The topological approach has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time.  So, while the
actions themselves may be incremental (i.e. unnecessary tasks will not be performed),
the pre-processing is not. We may try to incrementalise the topological sort by
storing the topological order between build runs and assume it to
be correct, but if the build determines it is wrong, fix it up.

This requires a way to abort tasks that have failed due to out-of-date
dependencies. It is also not minimal in the sense that a task may start, do some
meaningful work, then abort. However, in the case of an \hs{Applicative} system,
that work is zero.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}) is to simply build dependencies when they are
requested. By combining that with a transient log of which keys have already
been built, you can obtain a minimal build system.

@@

\subsection{Determining out-of-date keys} \label{sec-out-of-date}

@@

The idea of a dirty bit is to have one piece of persistent
information per key, saying whether the key is
\emph{dirty} or \emph{clean}. After a build, all bits are set to clean. When the
next build starts anything that changed between the two states is marked
dirty.
% ; and by marking additional things dirty/clean the build system can track
% what needs to rebuild.
When reaching a key, if it and all its dependencies are
clean, the key does not need recomputing.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell it (and its dependencies) dirty
if the user modifies it.
When rebuilding,
if a cell only depends on clean cells it is skipped, otherwise it is rebuilt and
marked dirty itself.
\todo{SLPJ}{What does ``marked dirty itself'' mean?}

@@

\Make uses file modification times, and compares files to their
dependencies, which can be thought of as a dirty bit which is set when
a file is newer than its dependencies. The interesting property of
this dirty bit is that it is not under control of \Make; rather it is
existing file-system information that is repurposed. In particular,
rebuilding a file automatically clears its dirty bit, and
automatically sets the dirty bit of the nodes depending on it. One
thing \Make does require is that file timestamps only go forward in
time -- something that can be violated by backup software.

@@

\subsection{Summary} \label{sec-design-space}

### 2018-03-15 paper/5-implementations.tex
Lots more from Simon

@@

The \Make build system builds files in a linear order based on a topological sort. For each file, it builds it if the output file is older than any of the inputs. We can express this in our framework with:

@@

We use the helper function \hs{reordering} which provides the key to build (\hs{k}), the dependencies of \hs{k} (\hs{dk}) and an action to build \hs{k} (\hs{act}). The dirty check is the same as \hs{makeDirtyBit}, with the added complication that because the \hs{Compute} is monadic it may not be possible to statically get a list of dependencies - if \hs{dk} is \hs{Nothing} then the action uses the monadic power and we conservatively rebuild anyway. If the check determines that the key is dirty, and needs rebuilding, we run \hs{act}, giving two possibilities:

>>>>

todo{SLPJ}{I think this entire section, including the mysterious M, can be deleted.  Or is
there anything that remains?}

@@

The \Make build system builds files in a linear order based on a topological sort.
(See \S\ref{sec-concurrency} for parallel \Make.)
For each file, it builds it if the output file is older than any of the inputs. We can express this in our framework with:

@@

We use the helper function \hs{reordering} which provides the key to build (\hs{k}), the dependencies of \hs{k} (\hs{dk}) and an action to build \hs{k} (\hs{act}). The dirty check is the same as \hs{makeDirtyBit}, with the added complication that because the \hs{Task} is monadic it may not be possible to statically get a list of dependencies - if \hs{dk} is \hs{Nothing} then the action uses the monadic power and we conservatively rebuild anyway. If the check determines that the key is dirty, and needs rebuilding, we run \hs{act}, giving two possibilities:

### 2018-03-15 paper/6-engineering.tex
Lots more from Simon

\begin{enumerate}
\item The \hs{topological} function can instead retain the dependency graph, and whenever all the dependencies of a node are complete, the node itself can be started.
\item The \hs{reordering} function can be made parallel in a few ways, but the most direct is to have $n$ threads reading entries from the list of keys. As before, if a key requires something not yet built, it is added to the end -- the difference is that sometimes things will be moved back not because they are out of order but because of races with earlier nodes that had not yet finished. As a consequence, over successive runs, potentially racey dependencies will be separated, giving better parallelism over time.
\item The \hs{recursive} function can be made parallel by working more directly on the \hs{Depeneds} structure, which gives multiple dependency keys at a time (all those arising from \hs{Applicative} dependencies). All such dependencies can be executed simultaneously, with a slightly smarter check for whether something is already executing that also blocks until an existing execution has finished.
\end{enumerate}

@@

In our model we define \hs{Compute} as a function - when give the same inputs it will always produce the same outputs. Alas, the real-world is not as obliging. Some examples of impure \hs{Compute} functions include:

@@

One tempting way of modelling non-determinism is to enrich \hs{Compute} from \hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus} respectively - both provide the underlying class with zero and choice. However, that seems to encode a different type of non-determinism - a kind of dependency-level non-determinism when most rules really experience a value-level non-determinism.

>>>>

\todo{SLPJ}{Can we replace ``node'' with ``task''? Everywhere.}
\begin{enumerate}
\item The \hs{topological} function can instead retain the dependency graph, and whenever all the dependencies of a node are complete, the node itself can be started.
\item The \hs{reordering} function can be made parallel in a few ways, but the most direct is to have $n$ threads reading entries from the list of keys. As before, if a key requires something not yet built, it is added to the end -- the difference is that sometimes things will be moved back not because they are out of order but because of races with earlier nodes that had not yet finished. As a consequence, over successive runs, potentially racey dependencies will be separated, giving better parallelism over time.
\item The \hs{recursive} function can be made parallel by working more directly on the \hs{Depeneds} structure, which gives multiple dependency keys at a time (all those arising from \hs{Applicative} dependencies). All such dependencies can be executed simultaneously, with a slightly smarter check for whether something is already executing that also blocks until an existing execution has finished.
\end{enumerate}

@@

In our model we define \hs{Task} as a function - when give the same inputs it will always produce the same outputs. Alas, the real-world is not as obliging. Some examples of impure \hs{Task} functions include:

@@

One tempting way of modelling non-determinism is to enrich \hs{Task} from \hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus} respectively - both provide the underlying class with zero and choice. However, that seems to encode a different type of non-determinism - a kind of dependency-level non-determinism when most rules really experience a value-level non-determinism.

### 2018-03-15 paper/7-related.tex
Lots more from Simon

Taking our abstraction, it is possible to encode \Fabricate assuming that nodes like \cmd{gcc -c util.c} are keys, there is a linear dependency between each successive node, and that the OS-level tracing can be lifted back as a monadic \hs{Compute} function\footnote{In fact, \Shake has an execution mode that can model \Fabricate{}-like build systems -- see \hs{Development.Shake.Forward} in the \Shake library.}. However, in our pure model the mapping is not perfect as \hs{gcc} writes to arbitrary files whose locations are not known in advance.

@@

The \hs{Compute} definition of:

\begin{minted}{haskell}
type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}

>>>>

Taking our abstraction, it is possible to encode \Fabricate assuming that nodes like \cmd{gcc -c util.c} are keys, there is a linear dependency between each successive node, and that the OS-level tracing can be lifted back as a monadic \hs{Task} function\footnote{In fact, \Shake has an execution mode that can model \Fabricate{}-like build systems -- see \hs{Development.Shake.Forward} in the \Shake library.}. However, in our pure model the mapping is not perfect as \hs{gcc} writes to arbitrary files whose locations are not known in advance.

@@

The \hs{Task} definition of:

\begin{minted}{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
